---
title: "Problemsheet 6"
author: "Emiko Bell"
date: "11/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(psych)
library(ggplot2)
library(car)
library(tidyverse)
library(kableExtra)
library(languageR)
library(gtools)
library(plyr)
library(RColorBrewer)
library(lsr)
```

# 1. Data manipulation and visualization - ugh, not again!!!

```{r}
data(lexdec)
selection <- c("Subject", "RT", "NativeLanguage", "Frequency","SubjFreq")
data <- lexdec[selection]

head(data) %>%
  knitr::kable(digits = 2)
summary(data)
```
```{r}
data$FRQ <- quantcut(data$Frequency, q=seq(0,1,1/2),
                       labels = c("low", "high"))
head(data)
summary(data)
```
This script added a FREQ variable with the categories of high and low. The script divided the Frequency into 2 groups at the 50% quantile.

## 1.1. Independent (between-group) vs. repeated-measures design

## 1.2. Summarizing data for independent and repeated-measures

```{r}
aggregate(RT~Subject,data, mean)

(aggFRQ <- aggregate(RT~Subject+FRQ,data, mean))

(aggNL <- aggregate(RT ~ Subject + NativeLanguage, data, mean))
```

### 1.2.1. Loading functions from file

```{r}
source("./data/summarySE.R")

summarySE
```

### 1.2.2. Summarizing your final pre-plot table

```{r}
(table <- summarySE(data=aggFRQ, measurevar = "RT", groupvars = "FRQ"))

ggplot(table, aes(y=RT, x=FRQ, fill=FRQ)) + geom_bar(stat = "identity")+
  geom_errorbar(aes(ymin=RT-se,ymax=RT+se), width=.1, size=1)+
  coord_cartesian(ylim=c(4, 7))
```
```{r}
source("./data/normDataWithin.R")
source("./data/summarySEwithin.R")

(table <- summarySEwithin(data=aggFRQ, measurevar = "RT", withinvars = "FRQ"))
ggplot(table, aes(y=RT, x=FRQ, fill=FRQ)) + geom_bar(stat = "identity") + coord_cartesian(ylim=c(4, 7)) +
  geom_errorbar(aes(ymin=RT-se,ymax=RT+se), width=.1, size=1)
```
Now, please plot reaction times as a function of NativeLanguage.
```{r}
(table <- summarySE(data = aggNL, measurevar = "RT", groupvars = "NativeLanguage"))

ggplot(table, aes(y=RT, x=NativeLanguage, fill=NativeLanguage)) + geom_bar(stat = "identity")+
  geom_errorbar(aes(ymin=RT-se,ymax=RT+se), width=.1, size=1)+
  coord_cartesian(ylim=c(4, 7))
```
```{r}
(sumagg <- aggregate(RT ~ Subject + FRQ + NativeLanguage, data, mean))

(sumtable <- summarySEwithin(data = sumagg, measurevar = "RT", betweenvars = "NativeLanguage", withinvars = "FRQ"))

ggplot(sumtable, aes(y = RT, x = NativeLanguage, fill = FRQ)) +
  geom_col(position = "dodge")+
  geom_errorbar(aes(ymin = RT - se, ymax = RT + se), width = .1, size = 1, position = position_dodge(width = 0.9)) +
  scale_fill_brewer(palette="Set3") +
  coord_cartesian(ylim = c(4, 7))
```

# 2. The t-test

## 2.1. The t-test as a general linear model

```{r}
summary(lm(RT~FRQ, data = aggFRQ))$coefficients
t.test(RT~FRQ, data = aggFRQ)
```

## 2.2. Conducting t-tests.
```{r}
(tNL <- t.test(RT~NativeLanguage, data = aggNL))

(tFRQ <- t.test(RT~FRQ, data = aggFRQ, paired = TRUE))
```

##2.3. Calculating effect size

### 2.3.1. r

```{r}
(t <- tNL$statistic[[1]])
(df <- tNL$parameter[[1]])
(r <- sqrt(t^2/(t^2+df))) 

(t <- tFRQ$statistic[[1]])
(df <- tFRQ$parameter[[1]])
(r <- sqrt(t^2/(t^2+df)))
```

### 2.3.2. Cohen’s d

For independent measures:
```{r}
(SD1 <- sd(aggNL$RT[ which(aggNL$NativeLanguage == "English") ]))
(SD2 <- sd(aggNL$RT[ which(aggNL$NativeLanguage == "Other") ]))
(SDpooled <- sqrt(((SD1^2)+(SD2^2))/2))
(Mdiff <- tNL$estimate[[2]]-tNL$estimate[[1]])
Mdiff/SDpooled

#OR

cohensD(RT ~ NativeLanguage, data = aggNL, method = "unequal")
```

For independent measures (unequal sample size - appropriate for this data):
```{r}
(SD1 <- sd(aggNL$RT[ which(aggNL$NativeLanguage == "English") ]))
(SD2 <- sd(aggNL$RT[ which(aggNL$NativeLanguage == "Other") ]))
(SDpooled <- sqrt((((12-1)*(SD1^2))+((9-1)*(SD2^2)))/(12+9-2)))
(Mdiff <- tNL$estimate[[2]]-tNL$estimate[[1]])
Mdiff/SDpooled

#OR

cohensD(RT ~ NativeLanguage, data = aggNL)
```

For dependent measures:
```{r}
cor(aggFRQ$RT[ which(aggFRQ$FRQ == "high")], aggFRQ$RT[ which(aggFRQ$FRQ == "low") ])

(SD1 <- sd(aggFRQ$RT[ which(aggFRQ$FRQ == "high") ]))
(SD2 <- sd(aggFRQ$RT[ which(aggFRQ$FRQ == "low") ]))
(SDwithin <- sqrt((SD1^2)+(SD2^2)-(2*0.9369256*SD1*SD2)))
(Mdiff <- 0.1192309)
Mdiff/SDwithin

#OR

cohensD(RT ~ FRQ, data = aggFRQ, method = "paired")
```

## 2.4. Reporting t-tests

*For independent measures:*

An independent samples t-test was run to compare the reaction times of participants depending on their native language. The results were significant (*t*(11.63) = -2.37, *p* = 0.04, 95% CI [-0.30, -0.01], $\hat{d}$ = 1.14), indicating that participants who spoke English as their native language had a significantly faster response time. 

*For dependent measures:*

A paired samples t-test was run to compare the reaction times of participants depending on the level of word frequency. The results were significant (*t*(20) = 7.94, *p* < 0.05, 95% CI [0.15, 0.09], $\hat{d}$ = 1.73), indicating that participants had a significantly faster response time for high frequency words.