---
title: "Problem Sheet 1"
author: "Emiko Bell"
date: "12/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(magrittr)
```

# 1. R as a calculator


**a.**
```{r}
a <- 344
```

**b.**
```{r}
n <- 13.76
25 * n - a
```
**c.**
```{r}
x <- 14
x^3
```

# 2. Basic vector operations

```{r}
x <- c(2, 4, 8, 10)
y <- c(1,3,5,7,9,11)
```

**a.** 

* x+1 would add 1 to all of the values in the vector
* y*2 would multiply each value in the vector by 2
* x+y does not have the same number of values in the vector and y does not divide by the number of values of x
* x*y has the same issue as above

**b.**

* length(x) would return the number of values in x
* length(y) would return the number of values in y

**c.**

* sum(x>5) would return the number of values in x above 5
* sum(x[x>5]) would take the values in x above 5 and add them together. () would operate for functions, and [] for the position of the values in the vectors.

**d.**

* y[-3] takes away the third value of the vector y
* y[x] shows each value that corresponds to the number in x (e.g. 2 is the first value in x, and would show the second value, 3, in y)
* y[y >= 7] shows values in y that is equal to or greater than 7

**e.**

* c(x,y) combines x and y consecutively and makes them into a single vector
* ?plot shows help for using the plot function

# 3. Revolving door problem

**a.**
```{r}
door <- c(159, 214,199, 219, 200, 165, 179, 175, 144, 201)
min(door)
```
```{r}
max(door)
```
**b.**
```{r}
mean(door)
```
**c.**
```{r}
mean(door) - max(door)
mean(door) - min(door)
```

# 4. Candy

**a.**
```{r}
person <- c(1, 2, 3, 4, 5, 6)
weight <- c(73, 67, 61, 62, 75, 79)
height <- c(1.87, 1.78, 1.75, 1.70, 1.77, 1.81)
```

**b.**
```{r}
bmi <- weight/height^2
format(bmi, digits = 4)
```
**c.**
```{r}
mean(bmi[1:3])
mean(bmi[4:6])
```
**d.**
```{r}
data <- data.frame(person, height, weight, bmi)
data %>%
  knitr::kable(digits=2)
```
**e.**
```{r}
data[6, 4]
```
**f.**
```{r}
data$person[data$bmi>mean(bmi)]
```
This string is returning the people who have greater than average BMI.
