---
title: "Problem Sheet 2"
author: "Emiko Bell"
date: "13/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(knitr)
library(magrittr)
library(psych)
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(rstatix)
library(car)
library(MASS)
```

# 1. Getting data into R
```{r}
thedataset <- read.table(here("data/TheDataSet.csv"), sep = " ", header = TRUE)
head(thedataset,n = 5L) %>%
  knitr::kable(digits = 2)
```



# 2. Exploring data

**2.1.** To get a sense of your data, use the functions summary, str, table. What types are the variables? Guess what the IVs and DVs are?
```{r}
summary(thedataset)
str(thedataset)
table(thedataset$text, thedataset$is_lure)
```
IVs are most likely the text, key, group, old.
DVs are most likely rt, correct, and success.

**2.2.** Pretty great that you now have a better understanding of this data set, but in this session we will largely ignore most of it and just focus on the dependent variable rt.

**2.2.1.** Make a new data frame in which you only keep success trials and call it AccData, for “accurate data”.
```{r}
success <- thedataset["success"]
accdata <- subset(thedataset, success > 0)
#Remember, this is how you took out a category of a variable while leaving the other variables as-is
#Can also you use filter for big data set
#Another solution: AccData <- TheDataSet[TheDataSet$success==TRUE,]
```

**2.2.2** What is the difference between skew & kurtosis? What is the skew and the kurtosis of the reaction times? Report all major central tendency measures and the dispersion/range of the reaction time distribution. You can answer all of this in 2 lines of code, using 2 functions.

Skew is either positive or negative (tails going to left or right), and the kurtosis is how "sharp" the peak of the distribution is.
```{r}
describe(accdata) %>%
  knitr::kable(digits = 2)
```



# 3. First plotting attempts

**3.1.** Install and load ggplot2
Done

**3.2.** Plot reaction time

```{r}
ggplot(data = accdata, aes(x=is_lure, y= rt)) + geom_boxplot()

ggplot(data = accdata, aes(x=is_lure, y=rt, fill=is_lure))+
  stat_summary(fun = "mean", geom = "bar", position="dodge")+
  stat_summary(fun.data = mean_cl_normal, geom="errorbar",
               position = position_dodge(width=0.9), width=0.2)

#another option: labs(title = "Reaction times for each type of lure", x = "Lure", y = "Reaction time(s)")
```


It looks like fun.y no longer works:
https://stackoverflow.com/questions/61216963/fun-y-and-stat-summary-stopped-working-in-ggplot



# 4. Assumptions of parametric data


**4.1.** Assumption of normality

**4.1.1.** Plot the distribution
```{r}
(hits.rt <- ggplot(data = accdata, aes(rt)) + geom_histogram(aes(y=..density..), colour = "black", fill = "blue")) + labs(x = "Reaction Time", y = "Density")
```

**4.1.2.** Add a normally distributed curve as a comparison - pick the mean and SD from the data.
```{r}
hits.rt + stat_function(fun = dnorm, args = list(mean = mean(accdata$rt, na.rm=T), sd = sd(accdata$rt, na.rm=T))) + labs(x = "Reaction Time", y = "Density")
```

**4.1.3.**
Q-Q Plot
```{r}
ggplot(accdata, aes(sample = rt)) + stat_qq(colour = "dark gray") + stat_qq_line(colour = "dark blue")
```
The rt data does not necessarily conform to the normal curve -- it tapers off after a while.

**4.1.4.**
```{r}
shapiro.test(accdata$rt)
```
The rt data is not normally distributed.

**4.2. Homogeneity of Variance**
```{r}
leveneTest(accdata$rt, accdata$group)
```
The groups are homogeneous in variance.

**4.3. Violated Assumptions**

**4.3.1.**
**4.3.1.1.-3.**
```{r}
accdata$logrt <- log(accdata$rt)
accdata$sqrtrt <- sqrt(accdata$rt)
accdata$speed <- 1/rt
describe(accdata$speed)
```
The kurtosis is wild...is this considered normal? I think it's way above the bounds of "normality." Not sure why this is called speed, other than the data is plotted from the mean of 1 and would show a greater or slower speed.

**4.3.2**
```{r}
lambdalist <- boxcox(rt~group, data = accdata)
lambda <- lambdalist$x[which.max(lambdalist$y)]
```

