---
title: "Problemsheet 3"
author: "Emiko Bell"
date: "20/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(psych)
library(car)
library(ggplot2)
library(here)
```

# 1. Revision

**1.1. Getting data into R**
```{r}
thedataset <- read.table(here("data/TheDataSet.csv"), sep = " ", header = TRUE)

head(thedataset, 10) %>%
  knitr::kable(digits = 2)
```
**1.2. Exploring data**
```{r}
summary(thedataset)
```

# 2. Sampling, creating data, and aggregating

**2.1. Probability sampling**
```{r}
#To take the first 12 rows, I used the following code. Note that the "," denotes all columns.
x <- thedataset[c(1:12),]
#Then, to randomly sample ROWS from the above 12 rows, I used the following code.
sample_n(x, 6, replace = TRUE)

#Sampling with replacement means that the first result will not affect the second results (e.g. can get the same value multiple times). Without replacement means that there is some kind of connection between two values (i.e. get each value only one time).
sample_n(x, 6, replace = FALSE)

#Setting the same seed would return the same results every time.
set.seed(44)
sample_n(x, 6)

#Predefined probability: we can change the weight to each variable
sample(c("pass", "fail"), 10, replace = TRUE, prob = c(.75, .25))
```
**2.2 Generate non-random data**
```{r}
#Repeating vectors

rep(c("pass", "fail"), times = 5)

rep(c("pass", "fail"), times = 5, each = 2)

rep(1, 40*(1-.8))

#Sequences

seq(from = 1, to = 10, by = 1)

seq(from = 1, to=10, by = 0.5)
```

**2.3. Random data generation**

```{r}
#uniform distribution
runif(6)

#normal distribution
rnorm(6)

#Now generate 2 different vectors of 100 random numbers. One from a normal distribution with a mean of 0 and a standard deviation of 2, and one from a uniform distribution with a minimum of -3 and a maximums of 3. Quickly plot these two using hist or a combination of plot and density.

normal <- rnorm(100, mean = 0, sd = 2)

uniform <- runif(100, min = -3, max = 3)

# historgram and/or density plot

hist(normal)
hist(uniform)

```

**2.3. Creating a data set**
```{r}
set.seed(88)
sandbox <- data.frame(
  pokemon=factor(rep(c("Charizard",
                       "Pikachu",
                       "Eevee"), each=100)),
  power=c(rnorm(100, mean=60, sd=2),
          rnorm(100, mean=55, sd=2),
          rnorm(100, mean=50, sd=2)),
  friendliness=c(rnorm(100, mean=10, sd=2),
               rnorm(100, mean=15, sd=2),
               rnorm(100, mean=20, sd=2))
  )


summary(sandbox)
ggplot(sandbox, aes(x = power)) +
geom_density(colour = "dark blue", fill = "dark blue") + geom_density(aes(x = friendliness), colour = "light blue", fill = "light blue", alpha = 0.5)
# the more transparent density plots are the friendliness values and the full one are the power values
```

**2.4. Aggregating**
```{r}
(aggPokemon <- aggregate(data = sandbox, FUN = mean, friendliness ~ pokemon))
```
If this makes it a bit more intuitive, you can read the arguments of the function as “From the data frame sandbox, aggregate the mean (this can be a different function, e.g. median, sd, sum) of the variable friendliness as a function of (~) the variable pokemon.

Now try to aggregate the mean of the power variable as a function of Pokémon type. You can also use cbind to aggregate both friendliness and power in the same data frame. Look at the Examples in the help function of aggregate for inspiration.

```{r}
(aggPokemon <- aggregate(data = sandbox, FUN = mean, power ~ pokemon))
```

# 3. Correlation

**3.1. Scatterplot**
```{r}
ggplot(sandbox, aes(x = power, y = friendliness)) + geom_point(colour = "dark green", shape = "triangle") + geom_smooth(method = lm, colour = "red", se = FALSE)
```

**3.2 From covariance to correlation**

**3.2.1. Covariance**
```{r}
cov(sandbox$friendliness, sandbox$power)
```

**3.2.2. Correlation**
```{r}
corrcoef <- cor(sandbox$friendliness, sandbox$power)

cor.test(sandbox$friendliness, sandbox$power)

corrcoef^2
```

**3.2.3. Non-parametric correlation**
```{r}
cor.test(sandbox$friendliness, sandbox$power, method = "spearman")
```

**3.2.4. Issues**

Charizard Test
```{r}
charizard <- filter(sandbox, sandbox$pokemon == "Charizard")

cor.test(charizard$friendliness, charizard$power)

ggplot(charizard, aes(x = power, y = friendliness)) + geom_point(colour = "red", shape = "circle") + geom_smooth(method = lm, colour = "firebrick4", se = FALSE)

```

Pikachu Test
```{r}
pikachu <- filter(sandbox, sandbox$pokemon == "Pikachu")

cor.test(pikachu$friendliness, pikachu$power)

ggplot(sandbox, aes(x = power, y = friendliness)) + geom_point(colour = "yellow", shape = "triangle") + geom_smooth(method = lm, colour = "gold4", se = FALSE)
```

Eevee Test
```{r}
eevee <- filter(sandbox, sandbox$pokemon == "Eevee")

cor.test(eevee$friendliness, eevee$power)

ggplot(sandbox, aes(x = power, y = friendliness)) + geom_point(colour = "cyan3", shape = "#") + geom_smooth(method = lm, colour = "blue4", se = FALSE)
```

# 4. Applying to The Data Set
```{r}
aggdata <- aggregate(data = thedataset, FUN = mean, cbind(success, rt) ~ code)

cor.test(aggdata$success, aggdata$rt)

ggplot(aggdata, aes(x = success, y = rt)) + geom_point(colour = "seagreen3", shape = "circle") + geom_smooth(method = lm, colour = "dodgerblue4", se = FALSE)

```
No significance, no relationship.
